---
title: "Is climate change real ?"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, include=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(lubridate)
library(here)
library(skimr)
library(janitor)
library(httr)
library(readxl)
library(vroom)
```



Following frequent claims from American politician that climate change was a hoax, we decided to do an unbiased investigation of this controversial topic.
Records show that global temperature is rising at an alarming pace, but is it only a short-term phenomenon, or a lasting trend ?
We will use our statistical tools and knowledge to assess global warming based on data published by NASA about the Combined Land-Surface Air and Sea-Surface Water Temperature Anomalies in the Northern Hemisphere.

The [tabular data of temperature anomalies used in this analysis can be found here](https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.txt). To define temperature anomalies we take the period between 1951-1980 as reference.

We first load the data from NASA website and store it as a dataframe.

```{r weather_data, cache=TRUE}

weather <- 
  read_csv("https://data.giss.nasa.gov/gistemp/tabledata_v3/NH.Ts+dSST.csv", 
           skip = 1, 
           na = "***")

```

We only keep the columns "Year" to "Dec" for the weather dataset, to have data for each year and month.
Then, we convert the dataframe to lond format using the `pivot_longer()` function.

```{r tidyweather}
weather_1 <- weather %>% 
  select(Year:Dec) 

tidyweather <- weather_1 %>% 
  pivot_longer(col=(Jan:Dec), names_to= "Month", values_to= "temp_deviation" )
```

## Plotting Information

To inspect the dataframe, we plot the data using a scatterplot and add a trendline.

```{r scatter_plot, fig1}

tidyweather <- tidyweather %>%
  mutate(date = ymd(paste(as.character(Year), Month, "1")),
         month = month(date, label=TRUE),
         year = year(date))

ggplot(tidyweather, aes(x=date, y = temp_deviation))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  labs (
    title = "Weather Anomalies", 
    y="Temperature Deviation",
    caption="Source: NASA's Goddard Institute for Space Studies"
  )+
  theme(
    axis.title.x = element_blank()
  )

```

```{r facet_plot_weather_anomalies, fig2}


ggplot(tidyweather, aes(x=date, y = temp_deviation))+
  geom_point()+
  geom_smooth(color="red") +
  theme_bw() +
  facet_wrap(~month)+
  labs (
    title = "Monthly Weather Anomalies",
    y="Temperature Deviation",
    caption="Source: NASA's Goddard Institute for Space Studies"
  )+
  theme(
    axis.title.x = element_blank()
  )

```

We remove data before 1800 and before using `filter`. Then, we use the `mutate` function to create a new variable `interval` which contains information on which period each observation belongs to. We can assign the different periods using `case_when()`.


```{r intervals}

comparison <- tidyweather %>% 
  filter(Year>= 1881) %>%     #remove years prior to 1881
  #create new variable 'interval', and assign values based on criteria below:
  mutate(interval = case_when(
    Year %in% c(1881:1920) ~ "1881-1920",
    Year %in% c(1921:1950) ~ "1921-1950",
    Year %in% c(1951:1980) ~ "1951-1980",
    Year %in% c(1981:2010) ~ "1981-2010",
    TRUE ~ "2011-present"
  ))

```


Now that we have the `interval` variable, we can create a density plot to study the distribution of monthly deviations (`delta`), grouped by the different time periods we are interested in. We set `fill` to `interval` to group and color the data by different time periods.

```{r, density_plot, fig3}

ggplot(comparison, aes(x=temp_deviation, fill=interval))+
  geom_density(alpha=0.2) +   #density plot with transparency set to 20%
  theme_bw() +                #theme
  labs (
    title = "Density Plot for Monthly Temperature Anomalies",
    y     = "Density", #changing y-axis label to sentence case
    x= "Temperature Deviation",
    fill= "Time Period" ,
    caption="Source: NASA's Goddard Institute for Space Studies"
  )

```

So far, we have been working with monthly anomalies. However, we might be interested in average annual anomalies. We can do this by using `group_by()` and `summarise()`, followed by a scatter plot to display the result. 

```{r averaging_annual_anomalies, fig4}

#creating yearly averages
average_annual_anomaly <- tidyweather %>% 
  group_by(Year) %>%   #grouping data by Year
  
  # creating summaries for mean delta 
  # use `na.rm=TRUE` to eliminate NA (not available) values 
  summarise(annual_average_delta = mean(temp_deviation, na.rm=TRUE)) 

#plotting the data:
ggplot(average_annual_anomaly, aes(x=Year, y= annual_average_delta))+
  geom_point()+
  
  #Fit the best fit line, using LOESS method
  geom_smooth() +
  
  #change to theme_bw() to have white background + black frame around plot
  theme_bw() +
  labs (
    title = "Average Yearly Anomaly",
    y     = "Average Annual Delta",
    caption="Source: NASA's Goddard Institute for Space Studies"
  )  +
  theme(
    axis.title.x = element_blank()
  )


```


## Confidence Interval for `delta`

[As NASA points out on their website](https://earthobservatory.nasa.gov/world-of-change/decadaltemp.php): 

> A one-degree global change is significant because it takes a vast amount of heat to warm all the oceans, atmosphere, and land by that much. In the past, a one- to two-degree drop was all it took to plunge the Earth into the Little Ice Age.

We will therefore construct a confidence interval for the average annual delta since 2011. 
To do so, we will try two different approaches, using a formula and using a bootstrap simulation with the `infer` package. 

```{r, calculate_CI_using_formula}
library(infer)

formula_ci <- comparison %>% 
  filter(interval=="2011-present") %>% 
  summarise(mean_delta= mean(temp_deviation, na.rm=TRUE), 
            sd_delta= sd(temp_deviation, na.rm=TRUE),
            count_delta= n(),
            se_delta= sd_delta/sqrt(count_delta),
            #calculating t_critical value for a 95% confidence interval
            t_critical= qt(0.975,   count_delta-1), 
            lower_delta= mean_delta- t_critical*se_delta,
            upper_delta= mean_delta+ t_critical*se_delta
            )

formula_ci
```


```{r, calculate_CI_using_bootstrap}

# use the infer package to construct a 95% CI for delta

boot_delta <- comparison%>%
  
  filter(interval=="2011-present") %>%
  
  # Specify the variable of interest
  specify(response = temp_deviation) %>%
  
  # Generate a bunch of bootstrap samples
  generate(reps = 1000, type = "bootstrap") %>%
  
  calculate(stat ="mean") 

percentile_ci <- boot_delta %>%
get_ci(level = 0.95, type = "percentile")

percentile_ci
```

Regardless of whether we choose the bootstrap method (we decided to do 1000 replications) or the CI formula method to estimate the population mean, we can say with 95% confidence that the mean temerature deviation has been around 1 degree since 2011.

Consequently, global warming is a real threat that we could prove using NASA data to construct confidence interval for the mean temperature deviation.
