---
title: "Physical activity benefits on US high schoolers health"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: show
---


```{r, setup, echo=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)

# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r load-libraries, echo=FALSE}
library(tidyverse)  # Load ggplot2, dplyr, and all the other tidyverse packages
library(mosaic)
library(ggthemes)
library(GGally)
library(readxl)
library(here)
library(skimr)
library(janitor)
library(broom)
library(lubridate)
library(tidyquant)
library(infer)
library(openintro)
```


We will analyse the [Youth Risk Behavior Surveillance System (YRBSS)](https://www.cdc.gov/healthyyouth/data/yrbs/index.htm) survey, released every two years by the Centers for Disease Control and Prevention conduct. This dataset contains several features about American high schoolers that will help us analyze health patterns.

First we have a look at the dataset to get a feel for missing values, summary statistics of numerical variables, and a very rough histogram. 

```{r, fig1}
data(yrbss)
glimpse(yrbss)
view(yrbss)
skim(yrbss)
```

## Exploratory Data Analysis

First we analyze the `weight` of participants in kilograms and find 1004 NA's. 

```{r, eda_on_weight, fig2}
#we analyse the dataframe
summary(yrbss$weight)

#we plot a graph to visualise the weight
yrb<-yrbss %>%
  filter(!is.na(weight))
yrb %>% 
  ggplot(aes(x=weight))+
  geom_boxplot()+
  labs (
    title = "Distribution of weight in high school students",
    subtitle= "Grades 9 to 12",
    caption="Source: Youth Risk Behavior Surveillance System, CDC",
    x= "Weight (Kg)"
  )+
  theme_clean()

#plotting the density
ggplot(yrb,aes(x=weight))+
  geom_density()+
  labs (
    title = "Distribution of weight in high school students",
    subtitle= "Grades 9 to 12",
    caption="Source: Youth Risk Behavior Surveillance System, CDC",
    x="Weight (Kg)",
    y="Density"
  )+
  theme_clean()



#calculating mean and median of weight 
mean_weight <- mean(yrb$weight)
median_weight <- median(yrb$weight)

paste("The mean of the high schoolers is:", mean_weight)
paste("The median of the high schoolers is: ", median_weight)
paste("Due to the mean, median and the density plot we can say that most high schoolers weigh around 64 kilograms ")

```

We now have a look at the relationship between a high schooler’s weight and their physical activity.

```{r, fig3}
summary(yrb$physically_active_7d)


#we convert the column yrb$physically_active_7d into a factor
yrb$physically_active_7d<-as.factor(yrb$physically_active_7d)


#boxplot showing relationship between high schooler’s weight and their physical activity
yrb %>% 
filter(!is.na(physically_active_7d)) %>% 
ggplot(aes(x=physically_active_7d,y=weight))+
  geom_boxplot()+
  labs (
    title = "Relationship between high schooler’s weight and their physical activity ",
    subtitle= "Grades 9 to 12",
    caption="Source: Youth Risk Behavior Surveillance System, CDC",
    y="Weight (Kg)",
    x="Physical Activity (days active per week)"
  )+
  theme_clean()

```


Create a new colun that will be `yes` if the highschoolers are physically active for at least 3 days a week, and `no` otherwise.

  
```{r, fig4}
#add column that calculates the students who are physically active for at least 3 days a week and the ones who arent 
yrbss <- yrbss %>% 
  mutate(physical_3plus = ifelse(physically_active_7d >= 3, "yes", "no"))



```

We now calculate a 95% confidence interval for the population proportion of high schools that are *NOT* active 3 or more days per week.

```{r, fig5}
#calculating CI 95% for the population proportion of high schools that are not active 3 or more days per week
stats<-yrbss %>% filter(!is.na(physical_3plus)) %>% 
  group_by(physical_3plus) %>% 
  summarise(count = n()) %>% 
  mutate(prop= count/sum(count))

se=sqrt(stats$prop[1]*(1-stats$prop[1])/stats$count[1])
ciupper=stats$prop[1]+1.96*se
cilower=stats$prop[1]-1.96*se

paste(" 95% confidence interval for the population proportion of high schools that are not active 3 or more days per week-", "Lower:" , cilower , " Upper: ", ciupper)
```

  
We now make a boxplot of `physical_3plus` vs. `weight`. We can see that there is a relationship between these two variables. We can tell that the high schoolers who excercise more are more likely due weigh more. Our assumption is the additional muscle weigh of those. 


```{r, boxplot, fig6}

#boxplot
yrbss %>% 
  filter(!is.na(physical_3plus)) %>% 
  ggplot(aes(x= physical_3plus, y= weight))+
  geom_boxplot()+
  theme_clean()+
  labs(
    title = "Students who exercise more frequently have a higher median weight!",
    subtitle= "Grades 9 to 12",
    caption="Source: Youth Risk Behavior Surveillance System, CDC",
    y="Weight (Kg)",
    x="Exercises more than 3 days a week?"
  )
 
```

## Confidence Interval

Boxplots show how the medians of the two distributions compare, but we can also compare the means of the distributions using either a confidence interval or a hypothesis test. 


```{r, fig7}
# we chose to compare the confidence intervals of both means or the ones excersing more and for the ones who excercise less
yrbss %>%
  group_by(physical_3plus) %>%
  filter(!is.na(physical_3plus)) %>% 
  summarise(mean_weight = mean(weight, na.rm = TRUE),
            sd_weight = sd(weight, na.rm=TRUE),
            count = n(),
            se_weight = sd_weight/sqrt(count),
            t_critical = qt(0.975, count-1), 
            margin_of_error = t_critical * se_weight,
            lower = mean_weight - t_critical * se_weight,
            upper = mean_weight + t_critical * se_weight
            )

```

There is an observed difference of about 1.77kg (68.44 - 66.67), and we notice that the two confidence intervals do not overlap. It seems that the difference is at least 95% statistically significant. Let us also conduct a hypothesis test.

## Hypothesis test with formula

Write the null and alternative hypotheses for testing whether mean weights are different for those who exercise at least times a week and those who don’t.

```{r, fig8}
#hyothesis testing for the whether mean weights are different for those who exercise at least times a week and those who don’t
t.test(weight ~ physical_3plus, data = yrbss)
```


## Hypothesis test with `infer`


Next, we will introduce a new function, `hypothesize`, that falls into the infer workflow. But first, we need to initialize the test, which we will save as `obs_diff`.

```{r}
obs_diff <- yrbss %>%
  specify(weight ~ physical_3plus) %>%
  calculate(stat = "diff in means", order = c("yes", "no"))

```

We now simulate the test on the null distribution, which we will save as null.


```{r}
null_dist <- yrbss %>%
  specify(weight ~ physical_3plus) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("yes", "no"))

```


We can visualize this null distribution with the following code:

```{r, fig9}
ggplot(data = null_dist, aes(x = stat)) +
  geom_histogram() +
  theme_clean()+
  labs(
    title = "Visualization of the null distribution",
    y="Count",
    x="Difference in means"
  )

```

Now that the test is initialized and the null distribution formed, we can visualize to see how many of these null permutations have a difference of at least `obs_stat` of `r obs_diff %>% pull() %>% round(2)`?

We can also calculate the p-value for your hypothesis test using the function `infer::get_p_value()`.

```{r, fig10}

null_dist %>% visualize() +
  shade_p_value(obs_stat = obs_diff, direction = "two-sided")

null_dist %>%
  get_p_value(obs_stat = obs_diff, direction = "two_sided")

```
